|

### 2025-09-06 manuscript/paper.Rmd
paper edits

<!-- @Rajski1961 -->
<!-- @McShane2019 -->
<!-- Csato 2020: The incentive (in)compatibility of group-based qualification systems. World cup qualifying has problems where teams are better off exerting lower effort -->

@@

There are hundreds of different competitions structures that dictate how a set of teams compete against each other to determine an overall ranking.  Common structures include bracket tournaments where if a team wins their competition they advance to the next round.  Bracket tournaments can be run as single elimination where competitors advance or are eliminated based on a single match (e.g. NCAA College Basketball Tournament, National Football League (NFL) playoffs, major tennis tournaments) or where each round is contested as a best of $k$ series (e.g. $k$ = 7: National Basketball Association (NBA) playoffs, National Hockey League (NHL) playoffs, $k$ varies by round: Major League Baseball (MLB), Women's National Basketball Association (WNBA)) and the series winner advance while the series loser is eliminated.  Additional styles of the bracket tournaments include variations where competitors are not eliminated after a single loss such as double elimination tournaments where competitors continue until they have lost twice.  Within double elimination tournament, there are several variations.  For instance, the NCAA College Wrestling tournament is a double elimination tournament, however, if a competitor loses a match they fall into the consolation bracket and the best they can finish after a loss before the finals is 3rd place making this not a "true" double elimination format.  Olympic judo uses a variant of this system with single elimination until the quarterfinals followed by a consolation bracket consisting of the four quarterfinal losers who still have a chance for a bronze medal^[in Olympic judo they award two bronze medals to the two  competitors who each win their side of the consolation bracket].

"True" double elimination - a bracket tournament where a competitor can still finish ranked first as long as a competitor has not lost two games - is used in NCAA baseball and softball for certain rounds in their postseason.  Additional variants of double elimination include double elimination brackets with a "true third" match which includes an additional match contested between the competitors ranked 2nd (loser of the finals) and 3rd (winner of the consolation bracket) at the end of the tournament *if they have not already competed against each other*. There is also a variation of double elimination where only those competitors who have lost to a competitor that makes the finals are placed into the consolation bracket and given the chance to compete for third place.  This style of competition is used in Olympic wrestling (both freestyle and Greco-Roman)^[As in Olympic judo, Olympic wrestling also award two bronze medals to the winners of each side of the consolation bracket].

@@

A natural question of interest after reviewing different types of competition structures is: which competition structure is best?  As with most complicated questions, the answer is "it depends".  One commonsense conceptualization for the best competition structure is that the final rankings produced by a competition structure should closely match the true rankings of the strength of the competitors. That is, a good competition structure should rank the true best team first, the true second best team second, and so on, until finally ranking the true worst team last.  We refer to this property of a competition structure as the efficacy of the structure (@Lasek2018-de, @Sziklai2022-kf).  If one were only interested in correctly ranking the true best competitor as first, with the ranks of the remaining competitors viewed as inconsequential, we refer to this here are effectivity (@Glenn1960-mn).  While there are many competition structures where efficacy and effectivity are useful measures^[efficacy is a useful measure for the EPL, for example, as the ranking of the teams matters at both the high end (i.e. entry into special tournaments) and the low end (i.e. relegation status) and effectivity is a useful measure for the NCAA College Basketball tournament (i.e. main goal is to crown a champion)], there are many other competitions where the correct ranking of more than one competitor, but not all the competitors is of import.  For instance, in Olympic competition, the goal of a competition structure is to correctly rank the top 3 competitors to award gold, silver, and bronze medals to^[Some sports award two bronze medals (e.g. wrestling, judo, boxing, tae kwon do)]. In many American sports, for instance the NFL, the goal of the competition structure of the regular season is to rank the top 7 teams in each conference to choose the playoff teams with the ranking of the teams 8 through 16 in each conference inconsequential in determining a Super Bowl champion^[The ranking of teams outside of the playoffs does matter for draft purposes though].  We coin the term "salient ranks" to refer to the ranks of a competition structure that matter.  That is the ranks 1, 2, and 3 in Olympic competition are the salient ranks in that competition structure.  In this notation, efficacy is a measure with with salient ranks 1 through $n$ where $n$ is the number of competitors, and effectivity is a measure where 1 is the only salient rank.  
<!-- Efficacy: @Lasek2018-de, @Sziklai2022-kf -->
<!-- Effectivity: @Glenn1960-mn -->

<!-- From Goosens 2024: The ability of a tournament to rank contestants according to their true strength is called efficacy (Lasek and Gagolewski, -->
<!-- 2018; Sziklai et al., 2022). Efficacy with respect to only the best player is also known as effectivity (Glenn, 1960) -->
<!-- or predictive power (Ryvkin and Ortmann, 2008; Vu and Shoham, 2011). -->

@@

Specifically, we use weighted mutual information with a weights based on squared error loss^[other loss functions can easily be incorporated into our method] between the ranks generated by a competition structure and the true ranks of the competitors.  Additionally, we include an indicator vector to specify the salient ranks, allowing us to compare different competition structures with different choices for salient ranks.

@@

<!-- $$ -->
<!-- max\{H(X), H(Y)\} = \sum_{r}  p(r) \log p(r) = n! \log n! -->
<!-- $$ -->

@@

What's interesting: 
1. Single and double are really close at 1 or 2, but they separate past 3.  
2.  The black curve crosses the yellow and green curve. 
3.  The yellow and black curves actually increases towards the end.

We examined the following structure: 
 - Single elimination, usual seeding (1 vs 8, 4 vs 5, 3 vs 6, 2 vs 7)
 - Double elimination

@@

![Zach add a caption \label{fig:fig1}](images/Normal_Information_Curve_Graph.png){width="408"}

Figure \ref{fig:fig1}

Based on the graph, the intuitive best tournament, round robin, does show to be the best. Double elimination is slightly better than single
elimination, with the gap growing wider as the top number of teams
looked at increases. However, those are under the assumption that the
true strengths were ordered correctly in the seeding. If there is a poor
seeding structure, as seen in the graph, it can drastically impact the
effectiveness of the tournament. In the extreme case seen, the
tournament can actually become worse than deciding games by random
chance (like flipping a coin). Another important discovery is that the
single elimination structure with equal strengths is better than every
possible permutation. The line for every possible permutation would
likely be similar to a round robin structure with all equal strengths,
so perhaps the closer the teams are in true strength, the closer round
robin and single elimination become.

>>>>

<!-- @Rajski1961 -->

<!-- @McShane2019 -->

<!-- Csato 2020: The incentive (in)compatibility of group-based qualification systems. World cup qualifying has problems where teams are better off exerting lower effort -->

@@

There are hundreds of different competitions structures that dictate how
a set of teams compete against each other to determine an overall
ranking. Common structures include bracket tournaments where if a team
wins their competition they advance to the next round. Bracket
tournaments can be run as single elimination where competitors advance
or are eliminated based on a single match (e.g. NCAA College Basketball
Tournament, National Football League (NFL) playoffs, major tennis
tournaments) or where each round is contested as a best of $k$ series
(e.g. $k$ = 7: National Basketball Association (NBA) playoffs, National
Hockey League (NHL) playoffs, $k$ varies by round: Major League Baseball
(MLB), Women's National Basketball Association (WNBA)) and the series
winner advance while the series loser is eliminated. Additional styles
of the bracket tournaments include variations where competitors are not
eliminated after a single loss such as double elimination tournaments
where competitors continue until they have lost twice. Within double
elimination tournament, there are several variations. For instance, the
NCAA College Wrestling tournament is a double elimination tournament,
however, if a competitor loses a match they fall into the consolation
bracket and the best they can finish after a loss before the finals is
3rd place making this not a "true" double elimination format. Olympic
judo uses a variant of this system with single elimination until the
quarterfinals followed by a consolation bracket consisting of the four
quarterfinal losers who still have a chance for a bronze medal[^1].

[^1]: in Olympic judo they award two bronze medals to the two
    competitors who each win their side of the consolation bracket

"True" double elimination - a bracket tournament where a competitor can
still finish ranked first as long as a competitor has not lost two
games - is used in NCAA baseball and softball for certain rounds in
their postseason. Additional variants of double elimination include
double elimination brackets with a "true third" match which includes an
additional match contested between the competitors ranked 2nd (loser of
the finals) and 3rd (winner of the consolation bracket) at the end of
the tournament *if they have not already competed against each other*.
There is also a variation of double elimination where only those
competitors who have lost to a competitor that makes the finals are
placed into the consolation bracket and given the chance to compete for
third place. This style of competition is used in Olympic wrestling
(both freestyle and Greco-Roman)[^2].

[^2]: As in Olympic judo, Olympic wrestling also award two bronze medals
    to the winners of each side of the consolation bracket

@@

A natural question of interest after reviewing different types of
competition structures is: which competition structure is best? As with
most complicated questions, the answer is "it depends". One commonsense
conceptualization for the best competition structure is that the final
rankings produced by a competition structure should closely match the
true rankings of the strength of the competitors. That is, a good
competition structure should rank the true best team first, the true
second best team second, and so on, until finally ranking the true worst
team last. We refer to this property of a competition structure as the
efficacy of the structure (@Lasek2018-de, @Sziklai2022-kf). If one were
only interested in correctly ranking the true best competitor as first,
with the ranks of the remaining competitors viewed as inconsequential,
we refer to this here are effectivity (@Glenn1960-mn). While there are
many competition structures where efficacy and effectivity are useful
measures[^3], there are many other competitions where the correct
ranking of more than one competitor, but not all the competitors is of
import. For instance, in Olympic competition, the goal of a competition
structure is to correctly rank the top 3 competitors to award gold,
silver, and bronze medals to[^4]. In many American sports, for instance
the NFL, the goal of the competition structure of the regular season is
to rank the top 7 teams in each conference to choose the playoff teams
with the ranking of the teams 8 through 16 in each conference
inconsequential in determining a Super Bowl champion[^5]. We coin the
term "salient ranks" to refer to the ranks of a competition structure
that matter. That is the ranks 1, 2, and 3 in Olympic competition are
the salient ranks in that competition structure. In this notation,
efficacy is a measure with with salient ranks 1 through $n$ where $n$ is
the number of competitors, and effectivity is a measure where 1 is the
only salient rank.\
<!-- Efficacy: @Lasek2018-de, @Sziklai2022-kf -->
<!-- Effectivity: @Glenn1960-mn -->

[^3]: efficacy is a useful measure for the EPL, for example, as the
    ranking of the teams matters at both the high end (i.e. entry into
    special tournaments) and the low end (i.e. relegation status) and
    effectivity is a useful measure for the NCAA College Basketball
    tournament (i.e. main goal is to crown a champion)

[^4]: Some sports award two bronze medals (e.g. wrestling, judo, boxing,
    tae kwon do)

[^5]: The ranking of teams outside of the playoffs does matter for draft
    purposes though

<!-- From Goosens 2024: The ability of a tournament to rank contestants according to their true strength is called efficacy (Lasek and Gagolewski, -->

<!-- 2018; Sziklai et al., 2022). Efficacy with respect to only the best player is also known as effectivity (Glenn, 1960) -->

<!-- or predictive power (Ryvkin and Ortmann, 2008; Vu and Shoham, 2011). -->

@@

Specifically, we use weighted mutual information with a weights based on
squared error loss[^6] between the ranks generated by a competition
structure and the true ranks of the competitors. Additionally, we
include an indicator vector to specify the salient ranks, allowing us to
compare different competition structures with different choices for
salient ranks.

[^6]: other loss functions can easily be incorporated into our method

@@

<!-- $$ -->

<!-- max\{H(X), H(Y)\} = \sum_{r}  p(r) \log p(r) = n! \log n! -->

<!-- $$ -->

@@

```{=html}
<!--

| Structure/Ranks | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 |
|:----------------------:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|:----:|
| Double Elimination | 0.4067 | 0.2517 | 0.1443 | 0.0916 | 0.0532 | 0.0390 | 0.0340 | 0.0290 |
| Single Elimination | 0.4023 | 0.2439 | 0.1338 | 0.0837 | 0.0478 | 0.0356 | 0.0312 | 0.0270 |
| Equal Strengths Single Elimination | 0.1297 | 0.0474 | 0.0224 | 0.0129 | 0.0072 | 0.0051 | 0.0050 | 0.0050 |
| Bad Seeding Single Elimination | 0.2679 | 0.0146 | 0.0102 | 0.0059 | 0.0032 | 0.0022 | 0.0020 | 0.0026 |
| Every Possible Permutation | 0.1036 | 0.0302 | 0.0113 | 0.0047 | 0.0018 | 0.0008 | 0.0011 | 0.0014 |
| Round Robin | 0.8851 | 0.8473 | 0.8096 | 0.7708 | 0.7337 | 0.7013 | 0.6755 | 0.6697 |
| Worst Possible | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| Best Possible | 1 | 1 | 1 | 1 | 1 | 1 | 1 | 1 |

-->
```

We examined the following structure:

-   Single elimination, usual seeding (1 vs 8, 4 vs 5, 3 vs 6, 2 vs 7)
    and normally distributed strengths

-   Single elimination, usual seeding (1 vs 8, 4 vs 5, 3 vs 6, 2 vs 7)
    and equal strengths (effectively each game is determined by a coin
    toss)

-   Single elimination, bad seeding (1 vs 2, 3 vs 4, 5 vs 6, 7 vs 8) and
    normally distributed strengths

-   Every Possible Permutation: One simulation for each $8! = 40320$
    possible combinations that final rankings may occur

-   Double elimination, usual seeding (1 vs 8, 4 vs 5, 3 vs 6, 2 vs 7)

-   Round Robin, normally distributed strengths

@@

![This chart compares the normalized mutual information (on log10 scale)
of various tournament structures against the number of top teams being
considered, where the team strengths are equally distanced based on a
normal distribution.
\label{fig:fig1}](images/Normal_Information_Curve_Graph.png){width="408"}

Based on the Figure 1 \ref{fig:fig1} , the intuitive best tournament,
round robin, does show to be the best. It is important to note that, due
to the weighting structure, the worst possible score approaches 0 as the
$i$ number of teams increases, but never reaches 0. So, the plot is
normalized so that the worst possible is 0 and the best possible is 1.
This explains the increase for a few of the lines at the end of the
plot. If there is a poor seeding structure, as seen in the graph, it can
drastically impact the effectiveness of the tournament. In the extreme
case seen, the tournament can actually become worse than deciding games
by random chance (like flipping a coin). Another important discovery is
that the single elimination structure with equal strengths is better
than every possible permutation. This is likely because, for a single
elimination structure, teams that play in the first round cannot both
advance to the next round in each simulation, leading to slightly more
information. Additionally, the bad seeding structure overlaps with the
equal strengths single elimination. This shows that, given that the best
team is the 1-seed, even a poorly structured single elimination
structure still results in the best team winning more than by random
chance. However, because the 2-seed plays the 1-seed in the first round,
the poor structure results in a less effective result because the top 2
can never be truly accurate according to the known strengths.

![Plot to further explore the difference between single elimination and
double elimination competition structures
\label{fig:fig2}](images/tournament_comparison_plot_se_de.png){width="409"}

Double elimination is slightly better than single elimination, with the
gap growing wider as the top number of teams looked at increases.
However, those are under the assumption that the true strengths were
ordered correctly in the seeding. While double elimination was expected
to be slightly better than single elimination, the increase in the gap
is likely due to noise that comes with how ties are handled. Ties led to
randomly assigning teams to a ranking, but with double elimination,
there were less ties that had to be dealt with, leading to more accurate
rankings.
